% !TEX root = ../Grunddatei.tex
\section{Introduction}

The systems programming language Rust has been on the rising edge of popularity \cite{stack_overflow_stack_2022}. Organizations like Discord \cite{jesse_howarth_why_2020}, Chromium \cite{dana_jansens_supporting_2023} or the Linux Kernel Organization \cite{noauthor_rust_nodate} are embracing the advantages of Rust and steadily integrating the language into their codebase.
The major motivation behind this transition are the promises Rust pledges regarding memory safety.
Memory safety is only one concern\footnote{Mind, a huge one: around 70 percent of the vulnerabilities found in Google Chrome and at Microsoft are related to memory bugs \cite{national_security_agency_software_2022}}, though, in the wide variety of safe and sound programs, e.g. for aviation.
Those programs need ways to reason about certain properties of their behaviour.
This is where formal verification tools, both for static and dynamic analysis, come into play.

Rust is a good starting point for designing formal correct programs.
It is fast, can switch between paradigms of a high and low-level programming language, and its strong type system and borrow-checker prevents memory safety issues and data races at compile time\footnote{While using safe Rust, that is}, without the need for additional tools and annotations like in C/C++ or SPARK.

Because it is a fairly new language, the ecosystem of formal verification tools in Rust isn't quite comparable (yet) to the languages just mentioned. There is a thriving community with the goal to make Rust a well-suited alternative, though:
for example, the Rust Verification Workshop \cite{peter_muller_eth_zurich_rust_2023}, whose third occurrence took place in April 2023, with 15 talks about various tools regarding verification in Rust. Another example is the Rust Formal Methods Interest Group \cite{rust_formal_methods_interest_group_welcome_nodate}, with its last meeting being held in October 2022. A third example would be from Google Research, though discontinued since September 2021, which provided a collection for tools and libraries called Rust Verification Tools (RVT) \cite{research_rust_2021}.

The goal of this thesis is to create an overview of the current state of verification in Rust and motivate projects and developers, which are in need of verification tools for their software, to consider moving to Rust as their day-to-day language of choice and thereby improving its popularity and distribution even more. It shall also show that Rust reduces the burden of verifying software, so that projects, which aren't particularly reliant on verification, might start using those tools regardless and improve their code quality by a big margin, while investing only a small amount of time and energy.

The thesis is split into two parts: in the first part, several verification projects are compared against each other, to give the reader a brief overview of the current state and properties of those. In the second part, I will choose a project and perform actual verification on Rust code, to demonstrate the tools capabilities and limitations.